Convective heat transport in stratified atmospheres at low and high Mach number

### 2017-03-03 tex/nusselt_paper.tex

At large Ma ($\epsilon = 0.5$), bulk thermodynamic structures are similar but
shock systems form in the upper atmosphere near downflow lanes 
(Fig. \ref{fig:entropy_snapshots}b,c) once Ra is sufficiently large.
Similar shock phenomena were reported in
both two \cite{cattaneo&all1990} and three \cite{malagoli&all1990} dimensional polytropic simulations previously.
As Ra is increased to large values 
(Fig. \ref{fig:entropy_snapshots}c), thermodynamic structures 
no longer span the whole domain but rather break up into 
small eddies which traverse the domain multiple
times before diffusing.

@@

The efficiency of convection is quantified by the Nusselt number (Nu).  
Nu is well-defined in RB convection
as the total flux normalized by the steady-state conductive flux 
\cite{johnston&doering2009, otero&all2002}.
In stratified convection Nu is more difficult to define, and we use
a modified version of a traditional stratified Nusselt number \cite{graham1975,hurlburt&all1984},
\begin{equation}
\text{Nu} \equiv \frac{\langle F_{\text{conv,z}} + F_{\text{cond,z}} - F_{\text{A}}\rangle}
{\langle F_{\text{cond,z}} - F_{\text{A}}\rangle} 
= 1 + \frac{\langle F_{\text{conv,z}}\rangle}{\langle F_{\text{cond,z}} - F_{\text{A}} \rangle}
\label{eqn:nusselt}
\end{equation}
where $F_{\text{conv,z}}$ and $F_{\text{cond,z}}$ are the 
z-components of $\bm{F}_{\text{conv}}$ and $\bm{F}_{\text{cond}}$,
respectively and $\langle \rangle$ are volume averages.  
$F_{\text{A}} \equiv -\langle\kappa\rangle \partial_z T_{\text{ad}}$ 
is the conductive flux of the proper corresponding adiabatic atmosphere
with $\partial_z T_{\text{ad}} \equiv - g / c_{P}$ 
for an ideal gas in hydrostatic equilibrium.  Here we specify
$\langle \kappa \rangle = \langle \rho\chi \rangle$, which is nearly
$\kappa_0$ when $\epsilon$ is small but can change appreciably for large
values of $\epsilon$.
In the Boussinesq approximation, under which $\grad S = 0$ only when 
$\grad T = 0$, this definition reduces the the traditional definition
of the Nusselt number \cite{otero&all2002, johnston&doering2009}.

The variation of Nu with Ra$_{\text{top}}$ is shown in Fig. \ref{fig:re_and_nu_v_ra}a.
At low to moderate Ra, Nu $\propto$ Ra$_{\text{top}}^{1/3}$ regardless of $\epsilon$ (NEED TO CITE A
RB PAPER WITH A 1/3 POWER LAW \cite{ahlers&all2009} maybe?).  At large Ra$_{\text{top}}$, Nu $\propto$ Ra$_{\text{top}}^{1/5}$ at $\epsilon = 0.5$.
The scaling of Nu with Ra is unclear at low $\epsilon$ (so I'll try to fill in that space with
data from comps...time to fish it out of Lou and do post-processing hell).

The Reynolds number and Peclet number,
\begin{equation}
\text{Re} = \frac{|\bm{u}| L_z}{\nu};\,\,\,\,\text{Pe} = \text{Pr}\,\text{Re},
\end{equation}
quantify the importance of advection to diffusion in the evolved
convective state.  For Pr = 1, such as in this work, Pe = Re.  
Our choice of $\{\nu,\chi\}\propto \rho_0^{-1}$ drastically changes
the value of Re between the top and bottom of the atmosphere.  We report values of
Re at the midplane ($z=L_z/2$) of the atmosphere in Fig. \ref{fig:re_and_nu_v_ra}b.
At low Ra$_{\text{top}}$, Re $\propto$ Ra$_{\text{top}}^{3/4}$, but at high Ra$_{\text{top}}$
where Ma $\geq 1$ for $\epsilon = 0.5$, this scaling gives way to a Re $\propto$ Ra$_{\text{top}}^{1/2}$.

@@

As these simulations evolve and the thermodynamic variables evolve, the density
profile evolves while remaining in hydrostatic equilibrium to zeroth order.  In 
Fig. \ref{fig:nrho_v_ra} we show the number of density scale heights present in the
evolved solution using two measures.  We find that once the average Ma of the domain
becomes approximately one, large density inversions begin to form in the boundary layers,
as was reported by \cite{brandenburg&all2005}.  The Regardless, the agreement of Nu
across $\epsilon$ despite the different effective stratifications felt in the simulations
is striking.

@@

@@

Furthermore, we have found that the stratification of these atmospheres evolves in a complex
manner, and future work should aim to understand the importance of changing stratification on
convective heat transport. Our studies
here will serve as a foundation both for understanding and 
comparing heat transport in stratified convection
to that in RB convection \cite{johnston&doering2009}, 
and for future studies of transport in stratified
convection,  both in determining the utility of simplified equation sets (such as the
anelastic equations) and in coming to understand more realistic systems, 
such as rapidly rotating atmospheres \cite{julien&all2012},
atmospheres bounded by stable regions \cite{hurlburt&all1986}, 
or regions with realistic profiles of $\kappa$.

>>>>

At large Ma ($\epsilon = 0.5$), bulk thermodynamic structures are similar but
shock systems form in the upper atmosphere near downflow lanes 
once Ra is sufficiently large  (Fig. \ref{fig:entropy_snapshots}b,c).
Similar shock phenomena were reported in
both two \cite{cattaneo&all1990} and three \cite{malagoli&all1990} dimensional polytropic simulations previously.
As Ra is increased to large values 
(Fig. \ref{fig:entropy_snapshots}c), thermodynamic structures 
no longer span the whole domain but rather break up into 
small eddies which traverse the domain multiple
times before diffusing.

@@

The efficiency of convection is quantified by the Nusselt number (Nu).  
Nu is well-defined in RB convection
as the total flux normalized by the steady-state conductive flux 
\cite{johnston&doering2009, otero&all2002}.
In stratified convection Nu is more difficult to define, and we use
a modified version of a traditional stratified Nusselt number \cite{graham1975,hurlburt&all1984},
\begin{equation}
\text{Nu} \equiv \frac{\langle F_{\text{conv,z}} + F_{\text{cond,z}} - F_{\text{A}}\rangle}
{\langle F_{\text{cond,z}} - F_{\text{A}}\rangle} 
= 1 + \frac{\langle F_{\text{conv,z}}\rangle}{\langle F_{\text{cond,z}} - F_{\text{A}} \rangle}
\label{eqn:nusselt}
\end{equation}
where $F_{\text{conv,z}}$ and $F_{\text{cond,z}}$ are the 
z-components of $\bm{F}_{\text{conv}}$ and $\bm{F}_{\text{cond}}$,
respectively and $\langle \rangle$ are volume averages.  
$F_{\text{A}} \equiv -\langle\kappa\rangle \partial_z T_{\text{ad}}$ 
is the conductive flux of the proper corresponding adiabatic atmosphere
with $\partial_z T_{\text{ad}} \equiv - g / c_{P}$ 
for an ideal gas in hydrostatic equilibrium.  Here we specify
$\langle \kappa \rangle = \langle \rho\chi \rangle$, which is nearly
$\kappa_0$ when $\epsilon$ is small but can change appreciably for large
values of $\epsilon$.
In the Boussinesq approximation, under which $\grad S = 0$ only when 
$\grad T = 0$, this definition reduces to the traditional definition
of the Nusselt number \cite{otero&all2002, johnston&doering2009}.

The variation of Nu with Ra$_{\text{top}}$ is shown in Fig. \ref{fig:re_and_nu_v_ra}a.
At low to moderate Ra$_{\text{top}}$, Nu $\propto$ Ra$_{\text{top}}^{1/3}$ regardless of $\epsilon$ (NEED TO CITE A
RB PAPER WITH A 1/3 POWER LAW \cite{ahlers&all2009} maybe?).  At large Ra$_{\text{top}}$, Nu $\propto$ Ra$_{\text{top}}^{1/5}$ at $\epsilon = 0.5$.
The scaling of Nu with Ra$_{\text{top}}$ is unclear at low $\epsilon$ (so I'll try to fill in that space with
data from comps...time to fish it out of Lou and do post-processing hell).

The Reynolds number and Peclet number,
\begin{equation}
\text{Re} = \frac{|\bm{u}| L_z}{\nu};\,\,\,\,\text{Pe} = \text{Pr}\,\text{Re},
\end{equation}
quantify the importance of advection to diffusion in the evolved
convective state.  For Pr = 1, such as in this work, Pe = Re.  
Our choice of $\{\nu,\chi\}\propto \rho_0^{-1}$ drastically changes
the value of Re between the top and bottom of the atmosphere.  We report values of
Re at the midplane ($z=L_z/2$) of the atmosphere in Fig. \ref{fig:re_and_nu_v_ra}b.
At low Ra$_{\text{top}}$, Re $\propto$ Ra$_{\text{top}}^{3/4}$, but at high Ra$_{\text{top}}$
where the average Ma $\approx 1$ for $\epsilon = 0.5$, this scaling gives way to a Re $\propto$ Ra$_{\text{top}}^{1/2}$.

@@

As the thermodynamic variables converge to their steady state values, the density
profile evolves while remaining in hydrostatic equilibrium to zeroth order.  In 
Fig. \ref{fig:nrho_v_ra} we show the number of density scale heights present in the
evolved solution using two measures.  We find that once the average Ma of the domain
becomes approximately one, large density inversions begin to form in the boundary layers,
as was reported by \cite{brandenburg&all2005}.  The Regardless, the agreement of Nu
across $\epsilon$ despite the different effective stratifications felt in the simulations
is striking.

@@

At low Ra and Ma, the scaling of Nu with Ra is reminiscent of RB convection.  However,
at high Ra and Ma, the scaling breaks down to a Ra$^{1/5}$ scaling unlike what is seen
in Boussinesq convection.  Similarly, the Reynolds number switches from a Ra$^{3/4}$ to
a Ra$^{1/2}$ scaling in this regime shift.  It is interesting that both of these
scaling laws turn over by roughly a factor of 2/3.

@@

Furthermore, we have found that the stratification of these atmospheres evolves in a complex
manner, and future work should aim to understand the importance of changing stratification on
convective heat transport. Our studies
here will serve as a foundation both for understanding and 
comparing heat transport in stratified convection
to that in RB convection \cite{johnston&doering2009}, 
and for future studies of transport in stratified
convection.  These results can be used to determine if simplified equation sets, 
such as the anelastic equations, carry heat in the same manner as the FC equations.
This work will also be useful in coming to understand more realistic systems, 
such as rapidly rotating atmospheres \cite{julien&all2012},
atmospheres bounded by stable regions \cite{hurlburt&all1986}, 
or regions with realistic profiles of $\kappa$.
